---
title: "05_eth_spi_ond2022_calc"
output: html_document
date: "2023-02-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## SPI

```{r cars}
library(rgee)
library(sf)
library(tidyverse)
library(here)
library(exactextractr) # for fast zonal stats (local)
library(lubridate)
ee_Initialize(drive = T)
sf::sf_use_s2(use_s2 = F) # too strict on topology
map(list.files(here("R"),full.names = T), ~source(.x))

write_raster <- c(T,F)[2]
```


```{r}
# load chirps
chirps_link <- "UCSB-CHG/CHIRPS/DAILY"
chirps_ic<- rgee::ee$ImageCollection(chirps_link)

# livelihood zone file path
lhz_fp <- "public/exploration/eth/ET_LHZ_2009/ET_LHZ_2009.shp"

# path to adm 3 layer to use
zip_path <- file.path(Sys.getenv("AA_DATA_DIR"),"public/raw/eth/cod_ab/eth_cod_ab.shp.zip")

# read in files
adm3 <- read_shape_zip(path = zip_path, layer = "eth_admbnda_adm3_csa_bofedb_2021")
lhz <- st_read(file.path(Sys.getenv("AA_DATA_DIR"),lhz_fp))

```


filter/wrangle down ROI.
- was initially trying to minimze ROI size for calculations. However, turns out ward level zonal stats are probably going to be too big/complex to do in GEE with `via="drive"` which is the simplest method for relatively big complex geoms. Therefore, it's better to do last step locally -- so these steps are bit pointless... nonetheless, they are fine to leave in as well
```{r}
lhz_of_interest <- lhz %>% 
    filter(LZTYPE %in% c("Agropastoral","Pastoral"))

lhz_of_interest_t <- lhz_of_interest %>% 
    st_transform(crs = st_crs(adm3))

roi_lhz_of_interest <- adm3[lhz_of_interest_t,] %>%
    st_transform(crs=4326)

roi_outline <-  roi_lhz_of_interest %>% 
    summarise()

roi_centroid <-  roi_outline %>% 
    st_centroid()


# for interactive GEE mapping
roi_centroid_ee <- sf_as_ee(roi_centroid)
```


# SPI

reduce chirps to OND SPI.

```{r}
ond2022_spi<- ee_date_range_anomaly(x = chirps_ic,time_range = c("2022-10-01","2022-12-31"))

# ended up getting icpac SPI - they used 1981-2010 as baseline but fit to gamma distribution. Curious to compare:

ond2022_spi_baseline2010<- ee_date_range_anomaly(x = chirps_ic,
                                    time_range = c("2022-10-01","2022-12-31"),
                                    baseline_years = c(1981:2010))


```


## visualize

The mapping/visualization can take a while to load - be patient. Not sure why sometimes it is lightning fast and other times sluggish
```{r}
Map$centerObject(eeObject = roi_centroid_ee,zoom = 6)
Map$addLayer(ond2022_spi,visParams = list(min= -0.27,
                                          max=0.477,
                                          palette=c("darkblue","green","red")
                                          ),
             name = "ond 2022spi")
```



## EE Zonal stats

**Please Skip Step** -- eval set to false - running here is just a waste of time

This failed - said task was too big, to do this you need to set up Google Cloud Storage bucket... could set this up or we just download the raster -- this might be better anyways to compare SPI calcs
```{r,eval=F}
median_ond2022_spi <- tidyrgee::ee_extract_tidy(x =ond2022_spi,
                                                y=roi_lhz_of_interest,
                                                stat="median",
                                                scale=5500, via="drive")
```


## Local zonal stats

let's convert to raster and download... the smaller we can make the clip outline the better (turns out we are going to get the whole country of ETH anyways so it doesn't matter)

Complex vector geometries are no bueno in GEE so let's simplify with bbox (this is still useful)

Create a bbox  & clip. 
```{r}
roi_bbox <- roi_lhz_of_interest %>% 
    st_bbox() %>% 
    st_as_sfc()

roi_bbox_ee <- sf_as_ee(roi_bbox)

on2022_spi_clipped <- ond2022_spi$clip(roi_bbox_ee)
on2022_spi_baseline2010_clipped <- ond2022_spi_baseline2010$clip(roi_bbox_ee)
```

convert to local raster 
```{r}
ond_2022_spi_c_r <- ee_as_raster(image = on2022_spi_clipped,region = roi_bbox_ee,scale = 5500)
ond_2022_spi_b2010_r <- ee_as_raster(image = on2022_spi_baseline2010_clipped,
                                     region = roi_bbox_ee,
                                     scale = 5500)

if(write_raster){
    raster::writeRaster(ond_2022_spi_c_r,"spi3_ond2022_eth_roi.tiff",overwrite=T)
    raster::writeRaster(ond_2022_spi_b2010_r,"spi3_ond2022_base2010_eth_roi.tiff",overwrite=T)
}
```

## zonal statistics

Let's make use of the speedy `{exactextractr} package!

```{r}

# this is so fast we can actually do for all admins
adm3_t <- st_transform(adm3,crs = st_crs(ond_2022_spi_c_r))


roi_ondspi3_median <- exact_extract(ond_2022_spi_c_r,
                                    adm3_t,
                                    append_cols=c("ADM1_EN","ADM2_EN","ADM2_PCODE","ADM3_EN","ADM3_PCODE"),
                                    'median')

roi_ondspi3_median <- roi_ondspi3_median %>% 
    rename(median_ond2022_spi="median") %>% 
    select(ADM3_PCODE,median_ond2022_spi)

roi_ondspi3_mean <- exact_extract(ond_2022_spi_c_r,
                                    adm3_t,
                                    append_cols=c("ADM1_EN","ADM2_EN","ADM2_PCODE","ADM3_EN","ADM3_PCODE"),
                                    'mean')

roi_ondspi3_mean <- roi_ondspi3_mean %>% 
    rename(mean_ond2022_spi="mean") %>% 
    select(ADM3_PCODE,mean_ond2022_spi)

adm3 %>% 
    left_join(roi_ondspi3_median,by="ADM3_PCODE") %>% 
    left_join(roi_ondspi3_mean, by="ADM3_PCODE") %>% 
    st_write("eth_climate.gpkg",layer = "adm3_ond2022_spi3")
```
