---
title: "05_eth_spi_ond2022_calc"
output: html_document
date: "2023-02-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## SPI

```{r cars}
library(rgee)
library(sf)
library(tidyverse)
library(here)
library(exactextractr) # for fast zonal stats (local)
ee_Initialize(drive = T)
sf::sf_use_s2(use_s2 = F) # too strict on topology
map(list.files(here("R"),full.names = T), ~source(.x))
```


```{r}
# load chirps
chirps_link <- "UCSB-CHG/CHIRPS/DAILY"
chirps_ic<- rgee::ee$ImageCollection(chirps_link)

# livelihood zone file path
lhz_fp <- "public/exploration/eth/ET_LHZ_2009/ET_LHZ_2009.shp"

# path to adm 3 layer to use
zip_path <- file.path(Sys.getenv("AA_DATA_DIR"),"public/raw/eth/cod_ab/eth_cod_ab.shp.zip")

# read in files
adm3 <- read_shape_zip(path = zip_path, layer = "eth_admbnda_adm3_csa_bofedb_2021")
lhz <- st_read(file.path(Sys.getenv("AA_DATA_DIR"),lhz_fp))
```


filter/wrangle down ROI.
```{r}
lhz_of_interest <- lhz %>% 
    filter(LZTYPE %in% c("Agropastoral","Pastoral"))

lhz_of_interest_t <- lhz_of_interest %>% 
    st_transform(crs = st_crs(roi))

roi_lhz_of_interest <- roi[lhz_of_interest_t,] %>%
    st_transform(crs=4326)

roi_outline <-  roi_lhz_of_interest %>% 
    summarise()

roi_centroid <-  roi_outline %>% 
    st_centroid()

roi_centroid_ee <- sf_as_ee(roi_centroid)
```


# SPI

reduce chirps to OND SPI
```{r}
ond2022_spi<- ee_date_range_anomaly(x = chirps_ic,time_range = c("2022-10-01","2022-12-31"))
ond2022_spi %>% ee_print()
```


## visualize

The mapping/visualization does take a while to load - be patient. Not sure why it seems faster in the code editor
```{r}
Map$centerObject(eeObject = roi_centroid_ee,zoom = 6)
Map$addLayer(ond2022_spi,visParams = list(min= -0.27,
                                          max=0.477,
                                          palette=c("darkblue","green","red")
                                          ),
             name = "ond 2022spi")
```



## EE Zonal stats
This failed - said task was too big, to do this you need to set up Google Cloud Storage bucket... could set this up or we just download the raster -- this might be better anyways to comapre SPI calcs
```{r,eval=F}
median_ond2022_spi <- tidyrgee::ee_extract_tidy(x =ond2022_spi,
                                                y=roi_lhz_of_interest,
                                                stat="median",
                                                scale=5500, via="drive")
```


## Local zonal stats
let's convert to raster and download... the smaller we can make the clip outline the better

Complex vector geometries are no bueno in GEE so let's simplify with bbox

Create a bbox  & clip. 
```{r}
roi_bbox <- roi_lhz_of_interest %>% 
    st_bbox() %>% 
    st_as_sfc()

roi_bbox_ee <- sf_as_ee(roi_bbox)

on2022_spi_clipped <- ond2022_spi$clip(roi_bbox_ee)
```

convert to raster 

```{r}
ond_2022_spi_c_r <- ee_as_raster(image = on2022_spi_clipped,region = roi_bbox_ee,scale = 5500)

raster::writeRaster(ond_2022_spi_c_r,"spi3_ond2022_eth_roi2.tiff",overwrite=T)
```

# zonal statistics

we need that `{exactextractr} package!

```{r}

# this is so fast we can actually do for all admins
adm3_t <- st_transform(adm3,crs = st_crs(ond_2022_spi_c_r))


roi_ondspi3_median <- exact_extract(ond_2022_spi_c_r,
                                    adm3_t,
                                    append_cols=c("ADM1_EN","ADM2_EN","ADM2_PCODE","ADM3_EN","ADM3_PCODE"),
                                    'median')

roi_ondspi3_median <- roi_ondspi3_median %>% 
    rename(median_ond2022_spi="median") %>% 
    select(ADM3_PCODE,median_ond2022_spi)

roi_ondspi3_mean <- exact_extract(ond_2022_spi_c_r,
                                    adm3_t,
                                    append_cols=c("ADM1_EN","ADM2_EN","ADM2_PCODE","ADM3_EN","ADM3_PCODE"),
                                    'mean')

roi_ondspi3_mean <- roi_ondspi3_mean %>% 
    rename(mean_ond2022_spi="mean") %>% 
    select(ADM3_PCODE,mean_ond2022_spi)

adm3 %>% 
    left_join(roi_ondspi3_median,by="ADM3_PCODE") %>% 
    left_join(roi_ondspi3_mean, by="ADM3_PCODE") %>% 
    st_write("eth_climate.gpkg",layer = "adm3_ond2022_spi3")
```
