---
title: "06_eth_icpac_spi_ond2022"
output: html_document
date: "2023-02-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# OND SPI 2022 Ethiopia

This script is used to analyze SPI data recieved from ICPAC.

It then takes the output of the NDVI analysis [here]() and merges the data into one output file

```{r setup}
library(rgee)
library(sf)
library(tidyverse)
library(here)
library(exactextractr) # for fast zonal stats (local)
library(lubridate)
map(list.files(here("R"),full.names = T), ~source(.x))

# path to adm 3 layer to use
zip_path <- file.path(Sys.getenv("AA_DATA_DIR"),"public/raw/eth/cod_ab/eth_cod_ab.shp.zip")
# read in adm3 COD
adm3 <- read_shape_zip(path = zip_path, layer = "eth_admbnda_adm3_csa_bofedb_2021")

# OND SPI raster
ond_spi <- terra::rast(file.path(Sys.getenv("AA_DATA_DIR"),"public/raw/eth/spi/spi_3_2022_dec.tif"))


# output of NDVI analysis
ndvi_stats_compiled_fp<- file.path(Sys.getenv("AA_DATA_DIR"),"public/exploration/eth/ndvi/eth_ndvi_woreda_stats_ond2022.csv")
ndvi_stats_compiled <- read_csv(ndvi_stats_compiled_fp)
```


Here I duplicated the OND raster and do some reclassification to simplify zonal statistics. By make anything greater than -2 NA and everything < -2 equal to one I can later just count the # of pixels =1 to get total pixels < - 2 SPI
```{r}
ond_reclass <-  ond_spi
ond_reclass[ond_reclass>-2] <- NA
ond_reclass[ond_reclass<=-2] <- 1
```

`exact_extract()` automatically does weighted counts based on % of pixel overlap with zonal boundary. First I:

- use the count statistic on the reclassified raster to just get count of values <= -2
- then I take the unclassified raster an calulate the following zonal stats per boundrary:
    + mean
    + median
    + count (in this case it is the count of all pixels per boundary)
    
```{r}

adm3_lt2 <- exact_extract(ond_reclass,
              st_transform(x = adm3,crs = st_crs(ond_spi)),
              append_cols=c("ADM1_EN","ADM2_EN","ADM2_PCODE","ADM3_EN","ADM3_PCODE"),
              # include_area=T,
              # coverage_fraction=T,
              c("count")
)

adm3_median_spi_area <- exact_extract(ond_spi,
              st_transform(x = adm3,crs = st_crs(ond_spi)),
              append_cols=c("ADM1_EN","ADM2_EN","ADM2_PCODE","ADM3_EN","ADM3_PCODE"),
              fun=c("mean","median","count")
)
```



Now that zonal stats have been run and data is in tabular format I can do some simple wrangling and calculations to get all the data how I need it for mapping.

```{r}
adm3_stats <- adm3_median_spi_area %>% 
    rename(total_pixels="count") %>% 
    left_join(
        adm3_lt2 %>% 
            rename(pixels_lte_neg2 = "count")
    ) %>% 
    mutate(
        pct_lte_neg2 = (round((pixels_lte_neg2/total_pixels)*100,1))
    )

adm3_stats_spatial <- adm3 %>% 
    left_join(
        adm3_stats 
    )
```

- clean up col names of SPI analysis
- clean up col names from NDVI analysis
- merge data sets and write file out.

```{r}

adm3_spi_stats_cleaned <- adm3_stats_spatial %>% 
    st_drop_geometry() %>% 
    select(matches("^ADM\\d_[EP]|*pix*|*pct*|*mean*|*median*")) %>% 
    select(ADM1_PCODE,ADM1_EN,ADM2_PCODE,ADM2_EN,ADM2_PCODE,ADM3_EN,ADM3_PCODE,
           `SPI mean` = mean, `SPI median`=median,`SPI % area <= -2` =pct_lte_neg2) %>% 
    mutate(across(matches("*mean*|*median*"),~round(.x,6)))




adm3_ndvi_stats_cleaned <- ndvi_stats_compiled %>% 
    select(
        ADM3_PCODE,
        `NDVI # pentads <= 80 % (mean)`= `number of pentads below 80% using mean`,
        `NDVI % pentads <= 80% (mean)`=`percent of pentads below 80%_mean`,
        `NDVI bin category (mean)` = `perc_binned_mean`,
        `NDVI # pentads <= 80% (median)`= `number of pentads below 80% using median`,
        `NDVI % pentads <= 80% (median)`= `percent of pentads below 80%_median` ,
        `NDVI bin category (median)` = `perc_binned_median`,
        `include (agropastoral/pastoral)`=include
        
    ) 

nrow(adm3_ndvi_stats_cleaned)
nrow(adm3_spi_stats_cleaned)

adm3_spi_stats_cleaned %>% 
    left_join(adm3_ndvi_stats_cleaned) %>% 
    write_csv(file.path(Sys.getenv("AA_DATA_DIR"),"public/exploration/eth/ndvi/eth_ndvi_spi_ond2022_adm3_stats.csv"),na = "")
```


- this file that I write to the geopacakge I use for mapping in QGIS.
```{r}
st_write(obj = adm3_stats_spatial,dsn = "eth_climate.gpkg",layer = "eth_adm3_ipac_spi_stats",append = T)
    
adm3_stats %>% 
    ggplot(aes(x= pct_lte_neg2))+
    geom_histogram()+
    scale_x_log10()

```